Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-10 wpe.tex
putting it in Phys Rev E Format
more PRE formatting
more pre formatting
first chunk of wpe section edit

\section{Measuring Structural Complexity {\color{red} NOT EDITABLE}}\label{sec:meaComplex}

@@

For the purposes of this paper, one can view entropy as a measure of complexity
and predictability in a time series.  A high-entropy time series is almost
completely unpredictable---and conversely.  This can be made more rigorous:
Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the
Shannon entropy rate is equal to the sum of the positive Lyapunov exponents,
$\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby
states of the system will diverge with time: $\left| \Delta x(t) \right| \approx
e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more
difficult prediction becomes.

Utilizing entropy as a measure of temporal complexity is by no means a new idea
\cite{Shannon1951, mantegna1994linguistic}.  Its effective usage requires
categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite
\emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is
effectively real-valued.  To get around this, one must discretize the data---
typically achieved by binning.  Unfortunately, this is rarely a good solution to
the problem, as the binning of the values introduces an additional dynamic on
top of the intrinsic dynamics whose entropy is desired.  The field of symbolic
dynamics studies how to discretize a time series in such a way that the
intrinsic behavior is not perverted, but these methods are fragile in the face
of noise and require further understanding of the underlying system, which
defeats the purpose of measuring the entropy in the first place.

@@

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_n$ as all
  $n!$ permutations $\pi$ of order $n$. For each $\pi \in \mathcal{S}_n$ we
  determine the relative frequency of that permutation occurring in $\{x_t\}_{t
  = 1,\dots,T}$:
  \begin{align*}
    p(\pi) = \frac{\left|\{t|t \leq T-n,\phi(x_{t+1},\dots,x_{t+n}) = \pi\}\right|}{T-n+1}
  \end{align*}
  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $n
  \ge 2$ is defined as
  \begin{align*}
  H(n) = - \sum_{\pi \in \mathcal{S}_n} p(\pi) \log_2 p(\pi)
  \end{align*}

@@

Notice that $0\le H(n) \le \log_2(n!)$ \cite{bandt2002per}.  With this in mind,
it is common in the literature to normalize permutation entropy as follows:
$\frac{H(n)}{\log_2(n!)}$.  With this convention, ``low'' entropy is close to 0
and ``high'' entropy is close to 1. Finally, it should be noted that the
permutation entropy has been shown to be identical to the Shannon entropy for
many large classes of systems \cite{amigo2012permutation}.

@@

%Liz commented this image out temporarily so that her mac doesn't hang
%when she scrolls past p12
    \includegraphics[width=\textwidth]{figs/SVD_wwpe}
    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
  \caption{The weighted permutation entropy of one run of
  SVD. The gray bands are regions where the window overlaps
  regimes. The window size used is 5,000 $\times$ 100,000 instructions
  and the word length is $6$. {\color{red} make sure this is actually updated to 6} For reference the instructions per cycle
  of \svd are plotted as a ghost behind this plot. Each color on the
  ghosted time series corresonds to the different regimes as selected
  by rapid shifts in WPE. From left to right each change in color
  represents a change in regime for 6 regimes in
  total.}\label{fig:wwpe}
\end{figure}

>>>>

\section{Measuring Structural Complexity }\label{sec:meaComplex}
{\color{red} NOT EDITABLE}

@@

For the purposes of this paper, one can view entropy as a measure of complexity and unpredictability in a time series.  A high-entropy time series is almost completely unpredictable---and conversely a low entropy signal is often quite predictable.  This can be made more rigorous: Pesin's relation \cite{pesin77} states that in chaotic dynamical systems, the Shannon entropy rate is equal to the sum of the positive Lyapunov exponents, $\lambda_i$. The Lyapunov exponents directly quantify the rate at which nearby states of the system will diverge with time: $\left| \Delta x(t) \right| \approx e^{\lambda t} \left| \Delta x(0) \right|$.  The faster the divergence, the more difficult prediction becomes.

Further we can consider the information contained in a single observable, say the \emph{present}, of the system, measured by the entropy. This information can be partitioned in to two components: the information shared with past observations --- the mutual information between the past and present --- and the information in the present which is not contained in the past --- the conditional entropy. The first part is known as \emph{redundancy}, as it is information in the present which is also in the past, and the second part is the \emph{entropy rate}. In theory, the more redundancy in a signal the more predictable it is. The difference between prediction methods is the types of redundancy (linear, non-linear, etc) they can detect and make use of.

Utilizing entropy as a measure of temporal complexity is by no means a new idea \cite{Shannon1951, mantegna1994linguistic}. Its effective usage requires categorical data: $x_t \in \mathcal{S}$ for some finite or countably infinite \emph{alphabet} $\mathcal{S}$, whereas data taken from real-world systems is effectively\footnote{Measurements from real devices are of finite precision, rendering them discrete. Unfortunately the space of values taken is so large that for the purposes of entropy calculations it can be taken as a continuum.} real-valued.  To get around this, one must discretize the data --- typically achieved by binning.  Unfortunately, this is rarely a good solution to the problem, as the binning of the values introduces an additional dynamic on top of the intrinsic dynamics whose entropy is desired\cite{bollt2001}. The field of symbolic dynamics studies how to discretize a time series in such a way that the intrinsic behavior is not perverted, but these methods are fragile in the face of noise and require further understanding of the underlying system, which defeats the purpose of measuring the entropy in the first place.

@@

  Given a time series $\{x_t\}_{t = 1,\dots,T}$. Define $\mathcal{S}_\ell$ as all $\ell!$ permutations $\pi$ of order $\ell$. For each $\pi \in \mathcal{S}_\ell$ we determine the relative frequency of that permutation occurring in $\{x_t\}_{t = 1,\dots,T}$:
  \begin{align*}
    p(\pi) = \frac{\left|\{t|t \leq T-\ell,\phi(x_{t+1},\dots,x_{t+\ell}) = \pi\}\right|}{T-\ell+1}
  \end{align*}
  Where $|\cdot|$ is set cardinality. The \emph{permutation entropy} of order $\ell \ge 2$ is defined as
  \begin{align*}
    H(\ell) = - \sum_{\pi \in \mathcal{S}_\ell} p(\pi) \log_2 p(\pi)
  \end{align*}

@@

Notice that $0\le H(\ell) \le \log_2(\ell!)$ \cite{bandt2002per}.  With this in mind, it is common in the literature to normalize permutation entropy as follows: $\frac{H(\ell)}{\log_2(\ell!)}$.  With this convention, ``low'' entropy is close to 0 and ``high'' entropy is close to 1. Finally, it should be noted that the permutation entropy has been shown to be identical to the Kolmolgorov-Sinai (KS) entropy for many large classes of systems \cite{amigo2012permutation}. In turn, and under very loose conditions, the KS entropy is equal to the Shannon entropy rate. What this normalization allows us to do, then, is approximate the redundancy of a signal by $1 - \frac{H(\ell)}{\log_2(\ell!)}$.

@@

%Liz commented this image out temporarily so that her mac doesn't hang
%when she scrolls past p12
    \includegraphics[width=\columnwidth]{figs/SVD_wwpe}
    %\caption{The MASE of ARIMA vs weighted permutation entropy. }
  \caption{The weighted permutation entropy of one run of
  SVD. The gray bands are regions where the window overlaps
  regimes. The window size used is 5,000 $\times$ 100,000 instructions
  and the word length is $6$. For reference the instructions per cycle
  of \svd are plotted as a ghost behind this plot. Each color on the
  ghosted time series corresonds to the different regimes as selected
  by rapid shifts in WPE. From left to right each change in color
  represents a change in regime for 6 regimes in
  total.}\label{fig:wwpe}
\end{figure}
